---
title: "IBM smoothing with INLA"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{IBM smoothing with INLA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

This vignette shows the INLA backend for integrated Brownian motion smoothing.

The global INLA model has one diffusion-precision parameter for the IBM transition density.  The locally adaptive model represents the log diffusion precision as a global log precision plus centered low-rank B-spline deviations that are shrunk toward zero, following the reduced-rank adaptive-smoothing-function idea of Yue, Simpson, Lindgren, and Rue (2014).  On the scaled model scale, each transition uses

\[
\begin{pmatrix}
f'_i - f'_{i-1} \\
f_i - f_{i-1} - h_i f'_{i-1}
\end{pmatrix}
\sim
N\left\{
\begin{pmatrix}0 \\ 0\end{pmatrix},
\lambda_i^{-1}
\begin{pmatrix}
h_i & h_i^2/2 \\
h_i^2/2 & h_i^3/3
\end{pmatrix}
\right\},
\]

where \(\lambda_i\) is the local diffusion precision.  In the adaptive INLA model,

\[
\log \lambda_i = \alpha + B_i^\top \beta,
\qquad
\beta_j \sim N(0, \sigma_\beta^2),
\]

where \(\alpha\) is the global log diffusion precision and the centered spline basis \(B_i\) represents departures from that global precision.  The argument `adaptive_deviation_sd` controls \(\sigma_\beta\). Smaller values make the adaptive model more reluctant to depart from the global IBM smoother. This is the same transition model used by the Stan backend, where the local process scale is \(\tau_i = 1/\sqrt{\lambda_i}\).  The INLA implementation computes both the precision matrix and the normalizing constant from this transition density.  The initial state is assigned the correlated unit-time IBM prior

\[
\begin{pmatrix} f_1 \\ f'_1 \end{pmatrix}
\sim
N\left\{
\begin{pmatrix}0 \\ 0\end{pmatrix},
\texttt{initial\_sd}^2
\begin{pmatrix}
1/3 & 1/2 \\
1/2 & 1
\end{pmatrix}
\right\}.
\]

The `initial_slope_sd` argument is retained only for backward compatibility and is ignored by the INLA backend.

## Global IBM

```{r global-inla, eval=FALSE}
library(ibmsmooth)

set.seed(1)
t <- sort(runif(40, 0, 10))
y <- sin(t) + rnorm(length(t), 0, 0.25)

fit_inla <- ibm(
  t = t,
  y = y,
  method = "inla",
  adaptive = FALSE,
  initial_sd = 5
)

plots <- plot(fit_inla)
plots$function_plot
plots$derivative_plot
summary(fit_inla)
```

## Locally adaptive IBM

```{r adaptive-inla, eval=FALSE}
t <- sort(runif(40, 0, 10))
y <- rnorm(length(t), 0, 0.25)
y <- y + ifelse(t < 5, 0, 2)

fit_inla <- ibm(
  t = t,
  y = y,
  method = "inla",
  adaptive = FALSE,
  initial_sd = 1
)

plots <- plot(fit_inla)
plots$function_plot
plots$derivative_plot
summary(fit_inla)

fit_inla_adapt <- ibm(
  t = t,
  y = y,
  method = "inla",
  adaptive = TRUE,
  n_knots = 10,
  adaptive_deviation_sd = 1.5,
  initial_sd = 1
)

plots <- plot(fit_inla_adapt)
plots$function_plot
plots$derivative_plot

precision_plot <- plot_precision_curve(fit_inla_adapt, n_samples = 1000)
precision_plot
```

The precision curve is evaluated at transition midpoints.  High process precision corresponds to stronger local smoothing; low process precision corresponds to a locally more variable derivative process.  The plotting helper uses posterior samples from `get_hyperparameter_samples()` and attaches the summarized plotting data to the returned object.

```{r precision-summary, eval=FALSE}
precision_summary <- attr(precision_plot, "precision_summary")
head(precision_summary)
```
