---
title: "Ordinary and locally adaptive IBM smoothing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ordinary and locally adaptive IBM smoothing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(ibmsmooth)
```

## Exact continuous-time model

The package uses the augmented state \(x(t)=(f'(t),f(t))^\mathsf T\). Over
\([t_j,t_{j+1}]\), with \(\Delta_j=t_{j+1}-t_j\),

\[
x_{j+1}\mid x_j,\lambda_j\sim N_2\{A_jx_j,\lambda_jQ_j\},
\quad
A_j=\begin{pmatrix}1&0\\\Delta_j&1\end{pmatrix},
\quad
Q_j=\begin{pmatrix}
\Delta_j&\Delta_j^2/2\\
\Delta_j^2/2&\Delta_j^3/3
\end{pmatrix}.
\]

This is an exact transition on irregular grids. It is not a finite-difference
approximation and does not replace observed time by operational time.

## Ordinary IBM

Ordinary IBM uses one diffusion standard deviation \(\tau\), so
\(\lambda_j=\tau^2\). `smoothing_prior` accepts valid Stan code involving the
positive parameter `tau`.

```{r ordinary, eval=FALSE}
fit_lognormal <- ibm(t, y)

fit_half_normal <- ibm(
  t, y,
  smoothing_prior = "tau ~ normal(0, 0.5);"
)

fit_half_t <- ibm(
  t, y,
  smoothing_prior = "tau ~ student_t(3, 0, 0.5);"
)
```

Because `tau` is constrained positive, symmetric Stan distributions in these
examples induce their corresponding truncated (half) distributions. Priors
are specified on the internally standardized response and time scale.

## Locally adaptive IBM

The adaptive model uses

\[
\sqrt{\lambda_j}=\gamma\xi_j,\qquad
\xi_j\sim C^+(0,1),\qquad
\gamma\sim C^+(0,s_\gamma).
\]

Most intervals can be strongly regularized while the heavy-tailed local scales
allow selected intervals to have substantially larger derivative innovations.
The derivative remains a continuous Brownian path; each local scale controls
how much its slope is permitted to evolve over an interval.

```{r adaptive, eval=FALSE}
fit_adaptive <- ibm(
  t, y,
  adaptive = TRUE,
  global_scale = 0.1,
  chains = 4,
  iter = 2000,
  adapt_delta = 0.99
)

# Same model in centered HMC coordinates
fit_adaptive_centered <- ibm(
  t, y,
  adaptive = TRUE,
  parameterization = "centered",
  global_scale = 0.1,
  chains = 4,
  iter = 2000,
  adapt_delta = 0.99
)

plot(fit_adaptive)
plot_diffusion(fit_adaptive)
summary(fit_adaptive)
```

Posterior samples for both the function and derivative are available directly:

```{r extraction, eval=FALSE}
f_draws <- get_samples(fit_adaptive, "f")
derivative_draws <- get_samples(fit_adaptive, "fprime")
lambda_draws <- get_samples(fit_adaptive, "lambda_interval")
```
