---
title: "Simulating and checking stochastic clocks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulating and checking stochastic clocks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(ibmsmooth)
```

This vignette gives short reproducible examples for checking the operational
time interpretation before fitting a model.

## Exact transition

```{r transition}
q <- 0.8
L <- ibm_clock_cholesky(q)
Q <- ibm_clock_covariance(q)

L %*% t(L)
Q
stopifnot(isTRUE(all.equal(L %*% t(L), Q)))
```

## Constant clock on an irregular grid

Equal rates do not imply equal operational-time increments when observed
interval lengths differ.

```{r irregular}
t <- c(0, 0.05, 0.4, 2.2, 2.5, 6)
sim_constant <- simulate_ibm_clock(t, lambda = 0.7, seed = 1)

data.frame(
  delta_t = diff(t),
  lambda = sim_constant$lambda_interval,
  q = sim_constant$q_interval
)
```

## Baseline acceleration

```{r acceleration}
lambda0 <- 0.5
r <- c(0, 0, 1.5, 1.5, 0)
sim_acceleration <- simulate_ibm_clock(
  t,
  lambda = lambda0 + r,
  seed = 2
)

data.frame(
  left = t[-length(t)],
  right = t[-1],
  lambda = sim_acceleration$lambda_interval,
  q = sim_acceleration$q_interval
)
```

## Persistent regimes and corners

Here the rate changes once and then persists. The latent function stays
continuous, while its observed-time derivative has distinct one-sided values
at the rate change.

```{r persistent}
lambda <- c(0.4, 0.4, 1.4, 1.4, 1.4)
sim_persistent <- simulate_ibm_clock(t, lambda = lambda, seed = 3)

knot <- 3L
c(
  left = sim_persistent$fprime_left[knot],
  right = sim_persistent$fprime_right[knot]
)

stopifnot(
  sim_persistent$fprime_left[knot] ==
    lambda[knot - 1L] * sim_persistent$B_operational[knot],
  sim_persistent$fprime_right[knot] ==
    lambda[knot] * sim_persistent$B_operational[knot]
)
```

These simulations are useful for prior predictive work: generate plausible
clock rates first, simulate the exact augmented IBM state, add an observation
model, and then fit the clock prior that expresses the intended regularization.
