---
title: "Simulating integrated Brownian motion"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulating integrated Brownian motion}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(ibmsmooth)
```

The state is ordered as \((f',f)\). For interval length \(\Delta\) and
derivative diffusion variance rate \(\lambda\), its innovation covariance is

\[
\lambda
\begin{pmatrix}
\Delta & \Delta^2/2\\
\Delta^2/2 & \Delta^3/3
\end{pmatrix}.
\]

The analytic Cholesky factor used by Stan can be checked directly.

```{r transition}
delta <- 0.8
lambda <- 0.4
L <- ibm_transition_cholesky(delta, lambda)
Q <- ibm_transition_covariance(delta, lambda)
stopifnot(isTRUE(all.equal(L %*% t(L), Q)))
Q
```

Constant `lambda` gives ordinary IBM. Supplying one value per interval gives
the locally varying diffusion used by the adaptive model.

```{r simulate}
t <- c(0, 0.1, 0.7, 2.2, 3)
ordinary <- simulate_ibm(t, lambda = 0.3, seed = 1)
adaptive <- simulate_ibm(t, lambda = c(0.02, 0.02, 1.2, 0.02), seed = 2)

plot(t, adaptive$f, type = "l", xlab = "t", ylab = "f(t)")
```
