---
title: "IBM smoothing with Stan"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{IBM smoothing with Stan}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

This vignette shows the Stan backend for integrated Brownian motion smoothing.

The Stan backend uses the same subordinated-IBM transition convention as the INLA backend.  On the scaled model scale, each transition uses

\[
\begin{pmatrix}
f'_i - f'_{i-1} \\
f_i - f_{i-1} - h_i f'_{i-1}
\end{pmatrix}
\sim
N\left\{
\begin{pmatrix}0 \\ 0\end{pmatrix},
\tau_i^2
\begin{pmatrix}
h_i & h_i^2/2 \\
h_i^2/2 & h_i^3/3
\end{pmatrix}
\right\}.
\]

INLA writes the same model with process precision \(\lambda_i = 1/\tau_i^2\).  The Stan and INLA backends also use the same correlated unit-time IBM prior for the initial state,

\[
\begin{pmatrix} f_1 \\ f'_1 \end{pmatrix}
\sim
N\left\{
\begin{pmatrix}0 \\ 0\end{pmatrix},
\texttt{initial\_sd}^2
\begin{pmatrix}
1/3 & 1/2 \\
1/2 & 1
\end{pmatrix}
\right\}.
\]

The adaptive Stan methods differ only in how they place a prior on the local process scales \(\tau_i\).  The INLA adaptive model instead writes \(\log \lambda_i = \alpha + B_i^\top\beta\), with \(\alpha\) as a global log precision and centered spline deviations \(\beta\) shrunk toward zero.


## Global IBM

```{r global-stan, eval=FALSE}
library(ibmsmooth)

set.seed(1)
t <- sort(runif(40, 0, 10))
y <- sin(t) + rnorm(length(t), 0, 0.25)

fit <- ibm(t, y, method = "stan", adaptive = FALSE, chains = 2, iter = 1000,
           initial_sd = 5)

plots <- plot(fit)
plots$function_plot
plots$derivative_plot

summary(fit)
```

## Locally adaptive IBM with a horseshoe prior

```{r adaptive-stan, eval=FALSE}
fit_hs <- ibm(
  t = t,
  y = y,
  method = "stan",
  adaptive = TRUE,
  stan_adaptive_method = "horseshoe",
  chains = 2,
  iter = 1000,
  initial_sd = 5
)

plots <- plot(fit_hs)
plots$function_plot
plots$derivative_plot

plot_precision_curve(fit_hs)
summary(fit_hs)
```
