---
title: "Stochastic-clock integrated Brownian motion"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Stochastic-clock integrated Brownian motion}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Model

The current `ibmsmooth` Stan models retain ordinary integrated Brownian motion
in operational time and adapt the rate at which operational time elapses. At
ordered grid points, the state is

\[
X_i = \{B(T_i), I(T_i)\}^{\mathsf T}, \qquad f(t_i)=I(T_i).
\]

For observed-time spacing \(\Delta_i=t_i-t_{i-1}\), an interval clock rate
\(\lambda_i\) gives operational-time increment

\[
q_i=\lambda_i\Delta_i.
\]

Conditional on \(q_i\), the package uses the exact transition

\[
X_i =
\begin{pmatrix}B_{i-1}\\I_{i-1}+q_iB_{i-1}\end{pmatrix}
+
\begin{pmatrix}
\sqrt{q_i}&0\\q_i^{3/2}/2&q_i^{3/2}/\sqrt{12}
\end{pmatrix}z_i,
\qquad z_i\sim N_2(0,I).
\]

Thus irregular observed spacing is handled through \(q_i\), rather than by
placing an exchangeable prior directly on operational-time increments.

## Example data

The package simulator follows the same transition used by Stan.

```{r data}
library(ibmsmooth)

set.seed(42)
t <- sort(c(0, runif(34, 0.1, 10), 10))
rate <- ifelse((t[-1] + t[-length(t)]) / 2 < 4, 0.5,
               ifelse((t[-1] + t[-length(t)]) / 2 < 7, 1.6, 0.7))
truth <- simulate_ibm_clock(t, lambda = rate, seed = 43)
y <- truth$f + rnorm(length(t), sd = 0.2)
```

## Constant clock

The nonadaptive model has \(\lambda_i=\lambda_0\) in every interval. It is
available as either `"ibm"` or `"nonadaptive"`.

```{r constant-fit, eval=FALSE}
fit_constant <- ibm_smooth(
  t, y,
  adaptive = "ibm",
  log_tau = list(mu = log(0.7), sd = 0.5),
  chains = 4,
  iter = 2000,
  adapt_delta = 0.95,
  seed = 101
)

plot(fit_constant)$function_plot
plot_clock(fit_constant, "lambda_interval")
summary(fit_constant)
```

## Baseline clock with sparse acceleration

The baseline-clock model uses

\[
\lambda_i=\lambda_0+r_i,\qquad r_i=\gamma\xi_i\geq0.
\]

It favors ordinary IBM at a common background rate and permits selected
intervals to accumulate operational time more rapidly.

```{r baseline-fit, eval=FALSE}
fit_baseline <- ibm_smooth(
  t, y,
  adaptive = "baseline_clock",
  log_tau = list(mu = log(0.5), sd = 0.5),
  zeta = 0.1,
  chains = 4,
  iter = 2000,
  adapt_delta = 0.99,
  seed = 102
)

plot(fit_baseline)$function_plot
plot_clock(fit_baseline, "lambda_interval")
plot_clock(fit_baseline, "r_interval")
```

When every \(r_i=0\), this model reduces exactly to the constant-clock model.

## Persistent clock with sparse changes

The persistent-clock model shrinks changes in adjacent log rates:

\[
\eta_i=\eta_{i-1}+\delta_i,\qquad
\eta_i=\log\lambda_i,
\qquad
\delta_i\mid\gamma,\xi_i\sim N(0,\gamma^2\xi_i^2).
\]

It favors long clock regimes and occasional transitions. The increments
\(\delta_i\) are not multiplied by \(\Delta_i\); interval length enters only
through \(q_i=\lambda_i\Delta_i\).

```{r persistent-fit, eval=FALSE}
fit_persistent <- ibm_smooth(
  t, y,
  adaptive = "persistent_clock",
  log_tau = list(mu = log(0.7), sd = 0.5),
  zeta = 0.1,
  chains = 4,
  iter = 2000,
  adapt_delta = 0.99,
  seed = 103
)

plot(fit_persistent)$function_plot
plot_clock(fit_persistent, "lambda_interval")
```

## Operational state and derivatives

Inside interval \(i\), the observed-time derivative is

\[
f'(t)=\lambda_iB(T(t)).
\]

At a rate jump, the left and right derivatives at a knot need not agree.
Extract them separately rather than requesting a globally continuous
derivative.

```{r extraction, eval=FALSE}
f_draws <- get_samples(fit_persistent, "f")
B_draws <- get_samples(fit_persistent, "B_operational")
left_derivative <- get_samples(fit_persistent, "fprime_left")
right_derivative <- get_samples(fit_persistent, "fprime_right")

# The unavailable outer derivative sides are returned as NA.
stopifnot(all(is.na(left_derivative[, 1])))
stopifnot(all(is.na(right_derivative[, ncol(right_derivative)])))
```

## Choosing an adaptive prior

The models encode different notions of simplicity:

* `"ibm"`: one clock rate throughout;
* `"baseline_clock"`: sparse positive departures from a global rate;
* `"persistent_clock"`: sparse changes from the preceding log rate.

The old curvature-shrinkage implementation remains available as
`"curvature_horseshoe"`. It has a distinct interpretation and should not be
treated as one of the stochastic-clock priors.

```{r legacy-fit, eval=FALSE}
fit_curvature <- ibm_smooth(
  t, y,
  adaptive = "curvature_horseshoe",
  chains = 4,
  iter = 2000,
  adapt_delta = 0.99
)
```

Always inspect divergences, effective sample sizes, and tree-depth saturation.
The baseline rate and adaptive scales can be weakly identified, so prior
predictive simulation and scientifically meaningful priors are preferable to
using sampler tuning alone.
